### 2018-07-19 submission/manuscript.Rmd
update manuscript from full read through

**Background.** A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, amplification, and sequencing steps. At each of these stages error can be introduced that will change the underlying bacterial community composition results. In this study, we focus on the amplification step's contribution to this overall error. To accomplish this we assessed 16S rRNA gene sequencing results in human fecal and mock community samples after using different high fidelity (HiFi) DNA polymerases and number of amplification cycles.

**Methods.** We extracted DNA from fecal samples (n=4) using a PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock communities (technical replicates n=4) consisting of previously isolated whole genomes of 8 different bacteria used the same approach. We first assessed GC dependent differences, error rate, sequence error prevalence, chimera prevalence, and correlation between chimera prevalence and number of Operational Taxonomic Units (OTUs) by polymerase and number of cycles. Next, differences in the number of OTUs was examined based on the polymerase and number of cycles used. Additionally, differences in the bacterial community composition by the Bray-Curtis index also was assessed based on polymerase and number of cycles. Based on these findings individual taxa differences based on polymerase and number of cycles was investigated. Finally, Random Forest models were created to test whether the bacterial community was better at classifying polymerases, number of cycles, or individual donor. We also assessed whether the most important taxa in the polymerase and number of cycle Random Forest models were also important in the model for individual donors.

**Results.** Predictably, we found noticeable differences in relative abundance based on high and low GC content (P-value $\leqslant$ `r round(max(filter(gc_data, bh < 0.05) %>% pull(bh)), digits = 2)`). Chimera prevalence in mock communities varied by polymerase with differences being most notable at 35 cycles (Kappa = `r round((filter(error_summary, cycles == "35x", taq == "K") %>% pull(chimera_prevalence_median))*100, digits = 2)`% (median) versus Platinum = `r round((filter(error_summary, cycles == "35x", taq == "PL") %>% pull(chimera_prevalence_median))*100, digits = 2)`%) and this variation persisted after chimera removal using VSEARCH. We also observed positive correlations between chimera prevalence and the number of OTUs with Platinum having the highest (R^2^ = `r max(pull(r2_table, r2))`) and Kappa having the worst (R^2^ = `r min(pull(r2_table, r2))`). When analyzing mock community samples the variation in the number of OTUs detected by the polymerases was observable as early as 20 cycles (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`). There also was a large range in the number of OTUs amplified by the polymerases at 35 cycles (Accuprime = `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion = `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`). When analyzing fecal samples we observed smaller differences in the number of OTUs. Additionally, the median number of OTUs only varied by HiFi DNA polymerase used at 35 cycles and not at 20 cycles like the mock community (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001). Random Forest models were most successful at classifying individual donor samples rather than polymerase or number of cycles used (P-value $\leqslant$ `r format(max(filter(dunn_rf_data, str_detect(var1, "fecal"), str_detect(var2, "fecal")) %>% pull(bh)), scientific = T, digits = 3)`). Additionally, the most important OTUs in the polymerase and number of cycle models were not the most important in the individual donor sample model.

**Conclusions.** Although there were 16S rRNA gene sequencing differences based on polymerase and number of cycles used, bacterial community composition differences are mostly only detectable in mock communities. Collectively, these results provide evidence that smaller biological differences between groups, based on 16S rRNA gene sequencing of fecal samples, can be consistently detected regardless of polymerase and number of cycles used.

@@

The bacterial community is reported to vary between case and control for a number of diseases [@Turnbaugh2008; @Baxter2016; @Sze2015; @Bonfili2017]. However, for diseases like obesity, the taxa identified have varied widely depending on the study [@Turnbaugh2008; @Zupancic2012]. Some of this variation could be due to error introduced during the 16S rRNA gene sequencing workflow. A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps. The preservation and extraction stages of the 16S rRNA gene sequencing workflow have been the most extensively studied [@contamination_Salter2014; @preservation_Song_2016; @storage_Bassis_2017; @review_Kim_2017]. For preservation and extraction stages of the workflow, it has been consistently found that there are errors in the observed bacterial community composition based on the kits used, but that these differences are smaller than the overall biological difference measured between samples with different kits [@preservation_Song_2016; @storage_Bassis_2017]. Since these studies use the same PCR approach while varying preservation or extraction method, the contribution of PCR error to this overall workflow is not as well characterized.

@@

Based on these observations, we created Random Forest models to classify samples based on number of cycles or polymerase used. When assessing the Random Forest models there was a difference between models built to classify number of cycles versus those built to classify polymerase used [P-value = `r format(filter(dunn_rf_data, str_detect(var1, "mock") & str_detect(var2, "mock")) %>% pull(bh), scientific = T, digits = 3)`]. Although there was a difference, neither model was successful at correctly classifying samples based on 16S rRNA gene sequencing data (cycles model, probability of correct classification = `r round(exp(-median(mock_cycles_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(mock_cycles_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(mock_cycles_rf_summary$logLoss)), digits = 2)`)(min - max) versus polymerase model, probability of correct classification = `r round(exp(-median(mock_poly_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(mock_poly_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(mock_poly_rf_summary$logLoss)), digits = 2)`)). Our observations in mock communities suggest that although differences are detected based on number of cycles or polymerase used, the overall community-wide differences are subtle.

***Few differences based on number of cycles and polymerase used were identified in fecal samples.***
Unlike mock community samples, an overall difference in the number of OTUs was only detected between polymerases at 35 cycles in fecal samples (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "F.value"]), digits = 2)`, P-value = `r format(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "bh"], digits = 2, nsmall = 2, scientific = T)`) [Table S4], and a Tukey post-hoc test failed to identify any differences between specific polymerases used (P-value `r ifelse(min(filter(fs_tukey_summary, sub_sample_level == 1000)[, "padj"]) > 0.05,  invisible(">"), invisible("check data"))` 0.05) [Table S5]. However, differences in the range of the number of OTUs detected for fecal samples was dependent on the polymerase used (e.g. Accuprime at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 4A]. There was also a trend for lower number of cycles (15-20) to result in a reduced range in the number of OTUs versus higher number of cycles (25, 30, and 35) for all polymerases (e.g. Phusion at 15 cycles = `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 4A]. Based on these observations, there are only small differences in the total number of OTUs detected in fecal samples that are based on number of cycles or polymerases used.

When using the Bray-Curtis index, we observed that polymerases and number of cycles had no affect on the bacterial community composition in fecal samples (P-value = `r round(filter(bray_fecal_combined, x1 == "polymerase") %>% select("Pr(>F)") %>% pull(), digits = 2)`). There was also no interaction between individuals, number of cycles, and polymerase used (P-value = `r round(filter(bray_fecal_combined, x1 == "interaction_ind_poly_cycle") %>% select("Pr(>F)") %>% pull(), digits = 2)`). When using PERMANOVA to test for community differences based on any of the number of cycles within polymerases, only Phusion had cycle dependent differences (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`). Additionally, Phusion was one of two polymerases that had enough sequences to be rarefied to 1000 sequences at 15 cycles. Within each respective 5-cycle increment comparison there was no difference in Bray-Curtis index between the polymerases (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05) [Figure 4B]. However, there was an overall decrease in Bray-Curtis index by cycle comparison group (e.g. the 15 cycle versus 20 cycle group compared to the 30 cycle versus 35 cycle group) (P-value `r ifelse(pull(fecal_kruskal_test_by_cycle_compare_summary, p.value) < 0.01, invisible("<"), invisible(">"))` 0.01) [Figure 4B]. Using a Dunn's post-hoc test the 15 cycle versus 20 cycle and 20 cycle versus 25 cycle comparison groups had a higher Bray-Curtis index than the 30 cycle versus 35 cycle comparison group (P-value `r ifelse(min(pull(dunns_test_by_cycle_compare_summary, P.adj)) < 0.05, invisible("<"), invisible(">"))` 0.05). In addition to assessing community composition by Bray-Curtis index, we also investigated differences in OTUs based on polymerase at each number of cycle used and no significant differences could be found (P-value $\geqslant$ `r round(min(pull(fecal_otu_tests, p.value)), digits = 2)`, corrected P-value = `r round(min(pull(fecal_otu_tests, bh)), digits = 2)`). Overall, these results suggest that in fecal samples the number of cycles and polymerase used only minimally change the bacterial community composition.

@@

The small differences observed due to error rate and chimera prevalence may be due to the actual DNA polymerase family being used within the PCR mix. DNA polymerases from distinct families have different binding affinities and error correction capacity [@polymerase_Ishino_2014]. However, based on the observations within our study this is unlikely to be the only contributor. Within our study the highest and lowest chimera prevalence both belonged to a family A polymerase (Platinum and Accuprime^TM^ respectively) [@polymerase_Ishino_2014]. Additionally, based on the information supplied by the respective manufacturers, the differences between the two PCR mixtures are not immediately apparent. Both PCR mixtures contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase and a platinum *Taq* antibody. Since it is not possible to know everything about the mixture beyond what was willingly provided by the manufacturer, it is possible that differences in how the recombinant *Taq* was generated or other compounds within the PCR mixture could be a contributing factor for the differences in error rate and chimera prevalence. Beyond the choice of the type of polymerase, there may be other ways to reduce the affect of polymerase dependent error rates and chimera prevalence on the downstream results.

Based on polymerase used, our observations generally found no difference when using the Bray-Curtis index to measure the bacterial community. Specifically, there was no difference in distance between successive 5-cycle increments within samples between the polymerases when using the Bray-Curtis index [Figure 4]. One possible reason for this outcome is that many of the OTUs generated by polymerase dependent error rates and chimera prevalence are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight these OTUs [@bc_index_Minchin1987]. The choice of downstream diversity metric could be an important consideration in helping to mitigate some of the observed polymerase dependent differences in error rate and chimera prevalence. Metrics that solely use presence/absence of OTUs (e.g. Jaccard [@Real1996], richness) may be less robust to polymerase dependent error rates and chimera prevalence. When choosing a distance metric, careful consideration of the biases introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account. One possible way to better choose polymerases may be based on the DNA polymerase family used in the PCR mixture.

@@

Although care should always be taken when choosing a polymerase for 16S rRNA gene sequencing, our observations show that the differences between a variety of polymerases are dwarfed by the actual biological variation in fecal communities between individuals. Therefore, if the biological signal of interest is similar to differences in fecal bacterial communities found between individuals, then the type of high fidelity DNA polymerase used will only minimally change the results.

@@

***Human and mock samples.***
Fecal samples were obtained from 4 individuals who were part of the Enterics Research Investigational Network (ERIN). The processing and storage of these samples were previously published [@erin_seekatz_2016]. Other than confirmation that none of these individuals had a *Clostridium difficle* infection, clinical data and other types of meta data were not utilized or accessed for this study. All samples were extracted using the MOBIO^TM^ PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA). The ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used for mock communities and was made up of *Pseudomonas aeruginosa*, *Escherichia coli*, *Salmonella enterica*, *Lactobacillus fermentum*, *Enterococcus faecalis*, *Staphylococcus aureus*, *Listeria monocytogenes*, and *Bacillus subtilis* at equal genomic DNA abundance (http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards).

***PCR protocol.***
The five different high fidelity DNA polymerases (hereto referred to as polymerases) that were tested included AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The polymerases activation time was 2 minutes, unless a different activation was specified by the manufacturer. The annealing and extension time for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md). For Kappa and Q5, the annealing and extension time also were from a previously published protocol [@taq_Gohl_2016]. For Phusion, the company defined activation and annealing times were used while the extension time followed the Accuprime and Platinum settings.

The number of cycles in the PCR for fecal and mock samples started at 15 and increased by 5 up to 35 cycles, with amplicons used at each 5-step increase for sequencing. The PCR of fecal DNA samples consisted of all 4 samples at 15, 20, 25, 30, and 35 cycles for each polymerase (total sample n=100). The mock communities had 4 replicates at 15, 20, 25, and 35 cycles and 10 replicates for 30 cycles for all polymerases (total samples n=130). No mock community sample had enough PCR product at 15 cycles for adequate 16S rRNA gene sequencing.

***Sequence processing.***
The mothur software program was used for all sequence processing steps [@mothur_schloss_2009]. The protocol has been previously published [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP). Two major differences from the published protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithm instead of average neighbor for OTU generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017]. Sequence error was determined using the 'seq.error' command on mock samples to compare back to the reference 16S sequences of *P. aeruginosa*, *E. coli*, *S. enterica*, *L. fermentum*, *E. faecalis*, *S. aureus*, *L. monocytogenes*, and *B. subtilis* [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013]. For simplicity, the use of the term 'error' within the context of this study moving forward encompasses changes in the downstream sequencing results due to the amplification step (e.g. higher GC content genomes, base substitutions, deletions, insertions, non-specific amplification, chimeras, etc.).

***Analysis workflow.***
To adjust for unequal sequencing, all samples were rarefied to 1000 sequences for downstream analysis. The analysis of the mock community of each polymerase for GC-based amplification differences, sequence error rate, number of sequences with an error, base substitution, and numbers of chimeras before and after chimera removal with VSEARCH was assessed. Additionally, the correlation between the number of chimeras and the number of OTUs was also assessed. The total number of OTUs, taxa differences, and Bray-Curtis indices were analyzed for both the fecal and mock community samples. Finally, Random Forest models were created to assess whether classification of polymerase, number of cycles, or individual could be performed best using 16S rRNA gene sequencing data. For the model used to classify individuals, all samples were included from all number of cycles and polymerases but only the individual labels were used. Both number of cycles and polymerase models included all samples but only the number of cycles or polymerase label was used for each respective model. Additionally, overlap between the most important OTUs to the three models was assessed using mean decrease in accuracy (MDA).

>>>>

**Background.** A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, amplification, and sequencing steps. At each of these stages error can be introduced that will change the underlying bacterial community composition results. In this study, we focus on the amplification step's contribution to this overall error. To accomplish this we assessed 16S rRNA gene sequencing results in mock community and human fecal samples after using different  amplification cycles and high fidelity DNA polymerases.

**Methods.** Amplification of mock communities (technical replicates n=4) consisting of previously isolated whole genomes of 8 different bacteria were amplifeid at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 high fidelity DNA polymerase. For fecal samples (n=4) we first extracted DNA using a PowerMag DNA extraction kit with a 10 minute bead beating step and then used the same approach as the mock community for the PCR amplification step. We first assessed GC dependent differences, error rate, sequence error prevalence, chimera prevalence, and correlation between chimera prevalence and number of Operational Taxonomic Units (OTUs) by polymerase and number of cycles. Next, differences in the number of OTUs was examined based on the polymerase and number of cycles used. Additionally, differences in the bacterial community composition was assessed based on polymerase and number of cycles using the Bray-Curtis index. Based on these findings individual taxa differences based on polymerase and number of cycles was also investigated. Finally, Random Forest models were created to test whether the bacterial community was better at classifying polymerases, number of cycles, or individual donor. We also assessed whether the most important taxa in the polymerase and number of cycle Random Forest models were also important in the model for individual donors.

**Results.** Predictably, we found noticeable differences in relative abundance based on high and low GC content (P-value $\leqslant$ `r round(max(filter(gc_data, bh < 0.05) %>% pull(bh)), digits = 2)`). Chimera prevalence in mock communities varied by polymerase with differences being most notable at 35 cycles (Kappa = `r round((filter(error_summary, cycles == "35x", taq == "K") %>% pull(chimera_prevalence_median))*100, digits = 2)`% (median) versus Platinum = `r round((filter(error_summary, cycles == "35x", taq == "PL") %>% pull(chimera_prevalence_median))*100, digits = 2)`%) and this variation persisted after chimera removal using VSEARCH. We also observed positive correlations between chimera prevalence and the number of OTUs with Platinum having the highest (R^2^ = `r max(pull(r2_table, r2))`) and Kappa having the worst (R^2^ = `r min(pull(r2_table, r2))`). When analyzing mock community samples the variation in the number of OTUs detected by the polymerases was observable as early as 20 cycles (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`). When analyzing fecal samples we observed smaller differences in the number of OTUs between polymerases. The median number of OTUs only varied by high fidelity DNA polymerase used at 35 cycles and not at 20 cycles like the mock community (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001). Random Forest models were most successful at classifying individual donor samples rather than polymerase or number of cycles used (P-value $\leqslant$ `r format(max(filter(dunn_rf_data, str_detect(var1, "fecal"), str_detect(var2, "fecal")) %>% pull(bh)), scientific = T, digits = 3)`). Additionally, the most important OTUs in the polymerase and number of cycle models were not the most important in the individual donor model.

**Conclusions.** Although there were 16S rRNA gene sequencing differences based on polymerase and number of cycles used, bacterial community composition differences were mostly only detectable in mock communities. Collectively, these results provide evidence that smaller biological differences between groups, based on 16S rRNA gene sequencing of human fecal samples, can be consistently detected regardless of polymerase and number of cycles used.

@@

The bacterial community is reported to vary between case and control for a number of diseases [@Turnbaugh2008; @Baxter2016; @Sze2015; @Bonfili2017]. However, for diseases like obesity, the taxa identified have varied widely depending on the study [@Turnbaugh2008; @Zupancic2012]. Although some of the differences could be attributed to study design, error introduced during the 16S rRNA gene sequencing workflow could also be another source of variation. A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps. The preservation and extraction stages of the 16S rRNA gene sequencing workflow have been the most extensively studied [@contamination_Salter2014; @preservation_Song_2016; @storage_Bassis_2017; @review_Kim_2017]. For preservation and extraction stages of the workflow, it has been consistently found that there are errors in the observed bacterial community composition based on the kits used, but that these differences are smaller than the overall biological difference measured between samples with different kits [@preservation_Song_2016; @storage_Bassis_2017]. Since these studies use the same PCR approach while varying preservation or extraction method, the contribution of PCR error to this overall workflow is not as well characterized.

@@

Based on these observations, we created Random Forest models to classify samples based on number of cycles or polymerase used. When assessing the Random Forest models there was a difference between models built to classify number of cycles versus those built to classify polymerase used [P-value = `r format(filter(dunn_rf_data, str_detect(var1, "mock") & str_detect(var2, "mock")) %>% pull(bh), scientific = T, digits = 3)`]. Although there was a difference, neither model was overly successful at correctly classifying samples based on 16S rRNA gene sequencing data (cycles model, probability of correct classification = `r round(exp(-median(mock_cycles_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(mock_cycles_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(mock_cycles_rf_summary$logLoss)), digits = 2)`)(min - max) versus polymerase model, probability of correct classification = `r round(exp(-median(mock_poly_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(mock_poly_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(mock_poly_rf_summary$logLoss)), digits = 2)`)). Our observations in mock communities suggest that although differences are detected based on number of cycles or polymerase used, the overall community-wide differences are not as large.

***Few differences based on number of cycles and polymerase used were identified in fecal samples.***
Unlike mock community samples, an overall difference in the number of OTUs was only detected between polymerases at 35 cycles in fecal samples (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "F.value"]), digits = 2)`, P-value = `r format(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "bh"], digits = 2, nsmall = 2, scientific = T)`) [Table S4], and a Tukey post-hoc test failed to identify any differences between specific polymerases used (P-value `r ifelse(min(filter(fs_tukey_summary, sub_sample_level == 1000)[, "padj"]) > 0.05,  invisible(">"), invisible("check data"))` 0.05) [Table S5]. However, differences in the range of the number of OTUs detected for fecal samples was dependent on the polymerase used (e.g. Accuprime at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 4A]. There was also a trend for lower number of cycles (15-20) to result in a reduced range in the number of OTUs versus higher number of cycles (25, 30, and 35) for all polymerases (e.g. Phusion at 15 cycles = `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 4A]. Based on these observations, there are small differences in the total number of OTUs detected in fecal samples that are based on number of cycles or polymerases used.

When using the Bray-Curtis index, we observed that polymerases and number of cycles had no affect on the bacterial community composition in fecal samples (P-value = `r round(filter(bray_fecal_combined, x1 == "polymerase") %>% select("Pr(>F)") %>% pull(), digits = 2)`). There was also no interaction between individuals, number of cycles, and polymerase used (P-value = `r round(filter(bray_fecal_combined, x1 == "interaction_ind_poly_cycle") %>% select("Pr(>F)") %>% pull(), digits = 2)`). When using PERMANOVA to test for community differences based on any of the number of cycles within polymerases, only Phusion had cycle dependent differences (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`). Additionally, Phusion was one of two polymerases that had enough sequences to be rarefied to 1000 sequences at 15 cycles. Within each respective 5-cycle increment comparison there was no difference in Bray-Curtis index between the polymerases (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05) [Figure 4B]. However, there was an overall decrease in Bray-Curtis index by cycle comparison group (e.g. the 15 cycle versus 20 cycle group compared to the 30 cycle versus 35 cycle group) (P-value `r ifelse(pull(fecal_kruskal_test_by_cycle_compare_summary, p.value) < 0.01, invisible("<"), invisible(">"))` 0.01) [Figure 4B]. Using a Dunn's post-hoc test the 15 cycle versus 20 cycle and 20 cycle versus 25 cycle comparison groups had a higher Bray-Curtis index than the 30 cycle versus 35 cycle comparison group (P-value `r ifelse(min(pull(dunns_test_by_cycle_compare_summary, P.adj)) < 0.05, invisible("<"), invisible(">"))` 0.05). In addition to assessing community composition by Bray-Curtis index, we also investigated differences in OTUs based on polymerase at each number of cycle used and no significant differences could be found (P-value $\geqslant$ `r round(min(pull(fecal_otu_tests, p.value)), digits = 2)`, corrected P-value = `r round(min(pull(fecal_otu_tests, bh)), digits = 2)`). Overall, these results suggest that in fecal samples the number of cycles and polymerase used minimally change the bacterial community composition.

@@

The small differences observed due to error rate and chimera prevalence may be due to the actual DNA polymerase family being used within the PCR mix. DNA polymerases from distinct families have different binding affinities and error correction capacity [@polymerase_Ishino_2014]. However, based on the observations within our study this is unlikely to be the only contributor. Within our study the highest and lowest chimera prevalence both belonged to a family A polymerase (Platinum and Accuprime^TM^ respectively) [@polymerase_Ishino_2014]. Additionally, based on the information supplied by the respective manufacturers, the differences between the two PCR mixtures are not immediately apparent. Both PCR mixtures contain a recombinant *Taq* DNA polymerase, a *Pyrococcus* spp GB-D polymerase, and a platinum *Taq* antibody. Since it is not possible to know everything about the mixture beyond what was willingly provided by the manufacturer, it is possible that differences in how the recombinant *Taq* was generated or other compounds within the PCR mixture could be a contributing factor for the differences in error rate and chimera prevalence. Beyond the choice of the type of polymerase, there may be other ways to reduce the affect of polymerase dependent error rates and chimera prevalence on the downstream results.

Based on polymerase used, our observations generally found no difference when using the Bray-Curtis index to measure the bacterial community. Specifically, there was no difference in distance between successive 5-cycle increments within samples between the polymerases when using the Bray-Curtis index [Figure 4]. One possible reason for this outcome is that many of the OTUs generated by polymerase dependent error rates and chimera prevalence are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight these OTUs [@bc_index_Minchin1987]. The choice of downstream diversity metric could be an important consideration in helping to mitigate some of the observed polymerase dependent differences in error rate and chimera prevalence. Metrics that solely use presence/absence of OTUs (e.g. Jaccard [@Real1996], richness) may be less robust to polymerase dependent error rates and chimera prevalence. When choosing a distance metric, careful consideration of the errors introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account.

@@

Although care should always be taken when choosing a polymerase for 16S rRNA gene sequencing, our observations show that the differences between a variety of polymerases are smaller than the actual biological variation in fecal communities between individuals. Therefore, if the biological signal of interest is similar to differences in fecal bacterial communities found between individuals, then the type of high fidelity DNA polymerase used will only minimally change the results.

@@

***Mock community and human samples.***
The ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used for mock communities and the bacterial component was made up of *Pseudomonas aeruginosa*, *Escherichia coli*, *Salmonella enterica*, *Lactobacillus fermentum*, *Enterococcus faecalis*, *Staphylococcus aureus*, *Listeria monocytogenes*, and *Bacillus subtilis* at equal genomic DNA abundance (http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards). The theoretical abundance of 0.12 used as a comparison point for our relative abundance was based on this equal genomic DNA composition of the mock community. Fecal samples were obtained from 4 individuals who were part of the Enterics Research Investigational Network (ERIN). The processing and storage of these samples were previously published [@erin_seekatz_2016]. Other than confirmation that none of these individuals had a *Clostridium difficle* infection, clinical data and other types of meta data were not utilized or accessed for this study. All samples were extracted using the MOBIO^TM^ PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA).

***PCR protocol.***
The five different high fidelity DNA polymerases that were tested included AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The polymerases activation time was 2 minutes, unless a different activation was specified by the manufacturer. The annealing and extension time for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md). For Kappa and Q5, the annealing and extension time also were from a previously published optimized protocol [@taq_Gohl_2016]. For Phusion, the company defined activation and annealing times were used while the extension time followed the Accuprime and Platinum settings.

The number of cycles in the PCR for mock and fecal samples started at 15 and increased by 5 up to 35 cycles, with amplicons used at each 5-step increase for sequencing. The mock communities had 4 replicates at 15, 20, 25, and 35 cycles and 10 replicates for 30 cycles for all polymerases (total samples n=130). No mock community sample had enough PCR product at 15 cycles for adequate 16S rRNA gene sequencing. The PCR of fecal DNA samples consisted of all 4 samples at 15, 20, 25, 30, and 35 cycles for each polymerase (total sample n=100).

***Sequence processing.***
The mothur software program was used for all sequence processing steps [@mothur_schloss_2009]. The protocol has been previously published [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP). Two major differences from the published protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithm instead of average neighbor for OTU generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017]. Sequence error was determined using the 'seq.error' command on mock samples to compare back to the reference 16S sequences of *P. aeruginosa*, *E. coli*, *S. enterica*, *L. fermentum*, *E. faecalis*, *S. aureus*, *L. monocytogenes*, and *B. subtilis* [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].

***Analysis workflow.***
To adjust for unequal sequencing, all samples were rarefied to 1000 sequences for downstream analysis. The analysis of the mock community of each polymerase for GC-based amplification differences, sequence error rate, number of sequences with an error, base substitution, and numbers of chimeras before and after chimera removal with VSEARCH was assessed. Additionally, the correlation between the number of chimeras and the number of OTUs was also assessed. The total number of OTUs, taxa differences, and Bray-Curtis indices were analyzed for both the fecal and mock community samples. Finally, Random Forest models were created to assess whether classification of polymerase, number of cycles, or individual could be performed best using 16S rRNA gene sequencing data. For the model used to classify individuals, all samples were included from all number of cycles and polymerases but only the individual label was used. Both number of cycles and polymerase models included all samples but only the number of cycles or polymerase label was used for each respective model. Additionally, overlap between the most important OTUs to the three models was assessed using mean decrease in accuracy (MDA).

***Statistical analysis.***
All analysis was done with the R (v `r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017]. Data transformation and graphing was completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) and colors selected using the viridis package (v `r paste(packageVersion("viridis")[1])`) [@tidyverse_2017; @viridis_citation_2017]. High and low GC content was determined based on the median GC percentage of either the V4 region or the average GC content of the whole genome of the bacterial species used in the mock community. Differences in the total number of OTUs were analyzed using an ANOVA with a tukey post-hoc test. For the comparison of the number of OTUS in fecal samples the data was normalized to each individual by cycle number to account for the biological variation. Bray-Curtis distance matrices were generated using mothur. The distance matrix data was analyzed using PERMANOVA with the vegan package (v `r paste(packageVersion("vegan")[1])`) [@vegan_citation] and Kruskal-Wallis tests within R. The Random Forest models were run using the caret package (v `r paste(packageVersion("caret")[1])`) [@caret_citation]. A total of 100 10-fold CV runs on different 80/20 splits of the data were run to generate a range of the Logloss value. The probability of a correct call was obtained from this Logloss value by taking the negative natural logarithm. For both error and chimera analysis, samples were tested using Kruskal-Wallis with a Dunns post-hoc test. Where applicable, correction for multiple comparison utilized the Benjamini-Hochberg method [@benjamini_controlling_1995].

***Reproducible methods.***
The code and analysis for this study can be found here https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017. The raw sequences can be found on the SRA at the following accession number SRP132931.
